Countering language attrition with PanLex and the Web of Data

### 2013-01-02 paper/2012/SWJ-PanLex/paper.tex

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{\small <http://panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{tbl:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

@@

\item For language and language varieties the classes
\texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes
\texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code} respectively.

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{\small plx:Definition} class, giving a textual representation
(\texttt{\small rdfs:label}) with more than just a lexeme in a certain
language variety (\texttt{\small plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the \texttt{\small plx:denotationMeaning} and \texttt{\small plx:denotationExpression} property respectively.
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
\item All approvers share the \texttt{\small plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{\small plx:registrationDate} pointing to an \texttt{\small xsd:date} literal,
\texttt{\small dc:title} to assign the title of a source,
\texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}

>>>>

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{\small <http://panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

@@

\item For language and language varieties the classes
\texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes
\texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}. % manual linebreak

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{\small plx:Definition} class, giving a textual representation
(\texttt{\small rdfs:label})
%with more than just a lexeme
in a certain
language variety (\texttt{\small plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{\small plx:Denotation} class pointing
to meanings and expressions via the properties \texttt{\small plx:denotationMeaning}
and \\ \texttt{\small plx:denotationExpression}. % manual linebreak
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
\item All approvers share the \texttt{\small plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{\small plx:registrationDate} pointing to an \texttt{\small xsd:date} literal,
\texttt{\small dc:title} to assign the title of a source,
\texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}
